I would like to express my sincere gratitude to my co-supervisors Dr. Bobby Stojanoski and Dr. Steven Livingstone for their guidance and mentorship throughout my time in the Master's program and during the completion of this thesis.
Their passion and dedication to their work was obvious from the moment I came onboard.
They gave thoughtful direction and valuable insights throughout my research.
I am especially thankful for the time they invested in me.
Because of them, I believe I have become a better researcher and have a deeper appreciation for how wonderful and complex the human brain is.

Their consistent availability and willingness to discuss ideas and give feedback was extremely valuable, and made working together feel easy and productive.
I'd also like to mention their belief in my potential was a huge factor in my development as a researcher, and I greatly appreciate their relentless encouragement to pursue further education, as well as the confidence they placed in my future.
Working with them has been a rewarding experience, and I am truly thankful for their support.

I'd like to thank my peers and friends, particularly those at the lab who volunteered their time to aid me in running my study, as well as those who participated in the study.
A special thanks to Dee for her support and encouragement, as well as teaching me how to make my diagrams look fantastic.

This thesis would not have been possible without my parents, Nila and Ronald, whose unwavering love and support was present throughout my Master's program.
They have always been there to listen, to motivate, and to offer steady guidance when I needed it most.
I am extremely fortunate to have had their encouragement and belief in me during every stage of this journey.
This would certainly not have been possible without them, and I am deeply grateful for everything they do and have done to help me achieve this goal. 
I would also like to acknowledge my younger brother Brandon who was one of the first test subjects to don the fNIRS cap and pilot the study.
I'm thankful for his help despite his workload as an engineering student, and his constant words of encouragement from the movie Interstellar: "It's not possible." "No, it's necessary". 

Lastly, I would also like to thank my grandparents for their steady support throughout my Master's program.
Their continued presence and care meant a great deal to me during this time.
I do not want to forget another important family member of mine, my dog Princess Leia, who sadly passed away.
She brought unlimited joy, happiness, and is a reminder of what truly matters in life.